---
title: "Tess_McCabe_lab4"
author: "Tess McCabe"
date: "2/12/2018"
output: html_document
---

```{r, echo=TRUE}

b <- read.table("data/pinecones.txt",header=TRUE)

str(b)
summary(b)
table(b$ring,b$tmt)           ## shows number of trees by ring and treatment

##histogram of cone counts in year 2000
hist(b$c00,breaks=0:92,probability=TRUE,ylim=c(0,1))

##plot of counts by tree size (x-axis) and CO2 (AMB is black, CO2 is red)
plot(b$diam,b$c00,col = b$tmt)


```


$$ L = \prod_{i=1}^{n} Pois \left( y_i \vert g(x_i) \right)$$
$$g(x_i) = a_0 x_i^2$$

```{r,echo=TRUE}
dia   <- b$diam       ## tree diameters
ncone <- b$c00       	## number of cones in 2000
cones <- (ncone > 0) 	## whether cones are present – this indicates whether tree is mature
tmt   <- b$tmt       	## treatment (elevated vs ambient)


# defining the negative log likelyhood of the model

##' @param a0   unknown parameter (slope of fecundity/size relationship)
##' @param x    tree diameter at breast height (cm)
##' @param y    cone counts (integer)
likFec = function(a0,x,y){
  g <- a0*x^2               ## expected mean fecundity, g(x),as a function of tree size
  -sum(dpois(y,g,log=TRUE)) ## negative log likelihood
}

out.fec <- optimize(likFec,c(0.001,1),x=dia,y=ncone)
out.fec
a0 <- out.fec$minimum  ## maximum likelihood parameter estimate

plot(dia,ncone)                    ## raw data scatter plot
dseq <- seq(1,30,length=300)       ## seq of x values for drawing curve
lines(dseq,a0*dseq^2,col=3,lwd=3)  ## add fecundity relationship
```

## Maturation Model

$$Bernoulli \left( m_i \vert \theta(x_i) \right)$$

```{r,echo=TRUE}
## example of probit
xseq = seq(0,30,length=200)
plot(xseq,pnorm(xseq,15,3),type='l')

```

**Task 1**

```{r, echo=TRUE}
b0 = 5
b1 = 5
param = c(b0,b1)
   
plot(dia, cones)
lines(dia, pnorm(dia, mean=b0, sd=b1), type="p", col="red")
    
```

**Task 2**

-  Similar to likFec and the models we constructed last week, code up the negative log likelihood for this model as an R function

```{r, echo=TRUE}
cones[cones==TRUE]<-1
cones[cones==FALSE]<-0

neg_log_mature<-function(param){
  b0 = param[1]
  b1 = param[2]
  theta<-pnorm(dia,mean=b0, sd=b1)# probit
  -sum(dbinom(cones,1, prob=theta, log = TRUE)) # Bernulli
}

neg_log_mature(param)
plot(-dbinom(cones,1, prob=theta, log = TRUE))
```

-  Test your function at your initial guess to verify that the function is working
Use a numerical optimization function to fit your log likelihood to the data. Make to use an optimization that supports multivariate problems (not optimize)

```{r, echo=TRUE}
vals<-optim(par=param, fn=neg_log_mature) # Produces NaN's but works. 
vals
```

##Combined Fecundity and Maturation

Probability of observing cones: 
$$p(y_i = 1,2,\dotsc) = Bernoulli \left( m = 1 \vert \theta (x_i) \right) Pois \left( y_i \vert g(x_i) \right) $$

Probability of observing no cones: 
$$p(y_i = 0) = Bernoulli \left( m = 0 \vert \theta (x_i) \right) + Bernoulli \left( m = 1 \vert \theta (x_i) \right) Pois \left( 0 \vert g(x_i) \right) $$
```{r, echo=TRUE}
##' @param param   vector of model parameters
##' @param dia     tree diameter at breast height (cm)
##' @param ncone   cone counts (integer)
likfit = function(param,dia,ncone){
  a0 = param[1]
  b0 = param[2]
  b1 = param[3]
  cones = ncone > 0

  ## trees with cones
  dia.cone   <- dia[cones > 0]              	    ## find just the trees with cones
  g.cone     <- a0 * dia.cone^2			              ## Fecundity fnc - g(x)
  theta.cone <- pnorm(dia.cone,b0,b1,log.p=TRUE) 	## maturation probit - theta(x)
  prob.cone  <- theta.cone + dpois(ncone[cones],g.cone,log=TRUE) ## lnL with cones
  
  ##trees with zero counts 
  dia.zero   <- dia[cones == 0]         ## find the trees without cones
  g.zero     <- a0 * dia.zero^2         ## Fecundity fnc - g(x)
  theta.zero <- pnorm(dia.zero,b0,b1)   	## maturation probit - theta(x)
  prob.zero  <- log((1-theta.zero) + theta.zero*dpois(0,g.zero)) ## lnL without cones

  return(-sum(prob.cone,prob.zero))     ## combined -lnL
}

param  <- c(0,0,0)        ## initial guess (update this based on the first two models!)
likfit(param,dia,ncone)   ## verify that the function works
out = optim(param,        ## run numerical MLE
             likfit,
             lower=c(0.001,10,1),
             upper=c(1,30,10),
             method="L-BFGS-B",dia=dia,ncone=ncone)
out                      ## display results
a0[1]  = out$par[1]         ## extract parameters for later use in plotting & analysis
b0[1]  = out$par[2]  # mean
b1[1]  = out$par[3]  # sd
lnL[1] = out$value
n[1]   = length(cones)

```

**Task 3**
- Make two plots showing fecundity (dia vs ncone) and maturation (dia vs cones) showing both the raw data and the best fit model. For the fecundity plot include two lines, one representing fecundity vs size given maturity, $g(x)$, and the second that is the expected number of cones observed, which combines fecundity and the probability of maturation $g(x) \cdot \theta(x)$

```{r, echo=TRUE}
 ### Dia vs cones

dseq <- seq(1,30,length=300)
plot(dia,ncone, xlab= "Diameter",ylab = "Fecundity" ) # Plot from before, fecundity given maturation

      
lines(dseq,a0*dseq^2,col=3,lwd=3) # Fecundity given maturity
lines(dseq,(a0*dseq^2)*pnorm(dseq, b0, b1),col=2,lwd=3) # Fecundity and the porbability of maturation

### Maturation

plot(dia, cones)
lines(dia,pnorm(dia, b0, b1), type= "p", col=2)


```

## Effects of Elevated CO2 on Reproduction

```{r, echo=TRUE}
## Ambient
dia.amb   = b$diam[tmt=="AMB"]   ## tree diameters
ncone.amb = b$c00[tmt=="AMB"]    ## number of cones

## Refit for just ambiant
out.amb = optim(param,likfit,method="L-BFGS-B",
             lower=c(0.001,10,1),upper=c(1,30,10),dia=dia.amb,ncone=ncone.amb)
a0[2]  = out.amb$par[1]
b0[2]  = out.amb$par[2]
b1[2]  = out.amb$par[3]
lnL[2] = out.amb$value
n[2]     = length(cones)
out.amb


```

**Lab Report Task 4**
- Repeat the analysis we just did but this time fit the model for just the elevated CO2 treatment (tmt == CO2). Save the output to the variable "out.elev" and store the same output values to the third element of the parameter vectors (e.g. a0[3]). Include a table of the values of the parameters in your lab report

```{r, echo=TRUE}
## Ambient
dia.CO2   = b$diam[tmt=="CO2"]   ## tree diameters
ncone.CO2 = b$c00[tmt=="CO2"]    ## number of cones

## Refit for just CO2iant
out.CO2 = optim(param,likfit,method="L-BFGS-B",
             lower=c(0.001,10,1),upper=c(1,30,10),dia=dia.CO2,ncone=ncone.CO2)
a0[3]  = out.CO2$par[1]
b0[3]  = out.CO2$par[2]
b1[3]  = out.CO2$par[3]
lnL[3] = out.CO2$value
n[2]     = length(cones)
out.CO2

table<-as.data.frame(cbind(a0,b0,b1))
rownames(table)<-c("whole dataset", "amb", "CO2")
table

```

## Model Selection

Likelihood ratio test: 
```{r, echo=TRUE}
dev.null <- 2*lnL[1]                  ## null model
dev.tmt  <- 2*lnL[2] + 2*lnL[3]       ## model with elevated vs ambient CO2 treatment
LR       <- dev.null - dev.tmt        ## Likelihood ratio
pval     <- 1-pchisq(LR,3)

LR
pval
```
**Task 5**

Create figures analogous to Task 3 that show the fits for the full model, the ambient-only fit, and the elevated-only fit (i.e. 3 lines per graph, with different colors or symbols for ambient vs elevated data). Based on these graphs, the parameter values, and the likelihood ratio test, answer the following questions:

```{r, echo=TRUE}

### Fecundity plots
dseq <- seq(1,30,length=300)
plot(dia,ncone, xlab= "Diameter",ylab = "Fecundity" ) # Plot from before, 
lines(dseq,(a0[1]*dseq^2)*pnorm(dseq, b0[1], b1[1]),col=2,lwd=3) 
lines(dseq,(a0[2]*dseq^2)*pnorm(dseq, b0[2], b1[2]),col=3,lwd=3) 
lines(dseq,(a0[3]*dseq^2)*pnorm(dseq, b0[3], b1[3]),col=4,lwd=3)
legend(x = "topleft", legend= c("Full datasat/ Null model", "Amb", "Co2" ), fill = c(2,3,4))

### Maturation plots
plot(dia, cones)
lines(dia,pnorm(dia, b0[1], b1[1]), type= "p", col=2)
lines(dia,pnorm(dia, b0[2], b1[2]), type= "p", col=3)
lines(dia,pnorm(dia, b0[3], b1[3]), type= "p", col=4)
legend(x = "topleft", legend= c("Full datasat/ Null model", "Amb", "Co2" ), fill = c(2,3,4))
```


1. Does elevated CO2 have an effect on fecundity? (Provide explanation for your answer).

According to this analysis, it does seem like elevated CO2 has some effect on the fecundity-diameter relationship. It looks like elevated CO2 increases the number of pine cones produced at most post 15 cm diameters. The ambient looks lower, and the p value was significant, suggesting that the visible difference is due to more than just chance. 


2. Is this effect simply due to there being larger trees in the elevated plots, or does CO2 affect the maturation rate and/or the fecundity of trees (i.e. the cone production at a given tree size)? If there is an effect, in what direction is it?

Looking at the maturation plot, there could be just an increase in maturation rate, but it would be a little one. The lines for CO2 and ambient are right on top of each other until about 15 cm. It could be argued that the 15 cm CO2 begins to diverge from the full model a tiny bit sooner than the ambient, but only a little. It looks like the affect is mostly in likelihood of producing a cone post 15 cm. 
