---
title: "Tess_McCabe_Lab2"
author: "Tess McCabe"
date: "1/29/2018"
output: html_document
---

```{r, echo= TRUE}
x = seq(-5,5,by=0.1)
p = seq(0,1,by=0.01) 
```

*1. Why does the height of the uniform PDF change as the width changes?*

Because a probability distribution function has to integrate to one. To keep the area under the uniform PDF at one, when the height increases, the length has to decrease. 

*2. What do the second and third arguments to the uniform specify? What are their default values?*

The second and third arguments are the min and max, which are by default set to 0 and 1 respectively. They have to be finite becuase otherwise the PDF would be a line infinitly close to the x axis.

*3) The Beta has a special case, Beta(1,1) that is equivalent to what other PDF?*

The Beta(1,1) is the same as the uniform distribution. 

```{r, echo=TRUE}
plot(dunif(p), type = "l") #the uniform
lines(dbeta(p,1,1)) #the special case of beta
#They look the same
```

*4) In the first panel, the mean is the same for each line (0.5).  What are the variances? (Hint: Calculate this analytically. Look up the distribution in one of the recommended references.)*

Beta equation: $f(x) = Γ(α + β) xα-1(1 – x)β-1 / (Γ(α) Γ(β))\  for\  0 ≤ x ≤ 1$
Varience of beta (According to the R help pages): $a*b/((a+b)^2 * (a+b+1))$

```{r, echo= TRUE}
a_1<-5
b_1<-5
varience_1<-(a_1*b_1)/((a_1+b_1)^2 *(a_1+b_1+1))
varience_1 # low varience

a_2<-1
b_2<-1
varience_2<-(a_2*b_2)/((a_2+b_2)^2 *(a_2+b_2+1))
varience_2 # medium varience

a_3<-0.2
b_3<-0.2
varience_3<-(a_3*b_3)/((a_3+b_3)^2 *(a_3+b_3+1))
varience_3 #high varience

```


*5) In the second panel, what are the means and medians of each of these curves?  (Hint: you'll need to calculate the mean analytically and use one of the variants of R's beta function to find the median.)*

The mean of beta is given by $a/(a+b)$. 

```{r, echo=TRUE}
betas<-c(6,4,2,1.25,1,0.5)
alphas<-rep(6,6)

means<-rep(NA,6)

for(i in 1:6){
  means[i]<- (alphas[i])/(alphas[i]+betas[i])
}
means

medians<-rep(NA,6)
for(i in 1:6){
  medians[i]<- qbeta(0.5, alphas[i], betas[i])
}
medians


```


*6) What are the arithmetric and geometric means of the three curves in the first panel? (Reminder: arithmetic means are means in the linear domain, geometric means are means in the log domain)*

$E(X) = exp(μ + 1/2 σ^2)$

```{r, ech0 =TRUE}

x <- 10^seq(-2,2,by=0.01)
plot(x,dlnorm(x,0),type='l',xlim=c(0,15),main="Changing the Mean")
lines(x,dlnorm(x,1),col=2)
lines(x,dlnorm(x,2),col=3)
legend("topright",legend=0:2,lty=1,col=1:3)

logmeans<-c(0,1,2) #log means

### Backing out arithmatic mean
arith_means<-rep(NA,3)
for(i in 1:3){
  arith_means[i]<- exp(logmeans[i] + 1/2 *1^2) #because sdlog is set to 1 by default
}
arith_means
```

*7) The last two panels compare a normal and a Laplace distribution with the same mean and variance.  How do the two compare?  In particular, compare the difference in the probabilities of extreme events in each distribution.*

The Laplace distribution doesn't converge to the x-axis as quickly as the normal distribution, so the distribution doesn't bias agains extream values as much as the normal distribution. The mean 

```{r, echo= TRUE}
```
